Testing Traversability of Protein Tunnels with Flexible Ligands Using Motion Planning

### 2017-06-15 paper.tex
sota planning, algoritmhs

@@

The widely used tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3} or 
a 3D grid~\cite{sehnal2013mole,petrek2006caver}.
These methods compute the tunnels assuming a spherical ligand (probe), and the resulting tunnels are represented as a sequence of spheres.
Bio-chemists then decide if such a tunnel can be used to transport a given ligand mainly based on the tunnel lengths and bottleneck (i.e.,
        the radius of the smallest sphere that can traverse the whole tunnel).
Typical ligands are however non-spherical and it is therefore difficult to estimate their traversability in tunnels computed for a spherical probe.
The decision based only on the spherical tunnels requires previour experties in the domain and yet it may be imprecise.

To provide chemists more insight into the behavior of non-spherical ligands, it is necessary to compute the trajecotries of the ligand considering it's shape and possibly also conformation changes, which can be formulated as a path planning problem in the a high-dimensional configuration space.

@@

This requires to compute trajectorie of the non-spherical ligand in the tunnels, which can be formulated as a path planning problem in a high-dimensional configuration space $\C$.
The configuration space is formed by all possible configurations of the ligand in the tunnel, i.e., considering its rotation, translation and possibly also other degrees of freedom responsible for conformation changes.
The dimension of the configuration space is given my the DOF of the ligand, i.e., 6D for a rigid and 6D + $n$ for a flexible
ligand with $n$ DOFs.
Sampling-based motion planning methods~\cite{Lav06} can be used to search this high-dimensional configuration space.
The idea of sampling-based motion planning is to randomly sample $\C$ and classify the samples as free or non-free using collision detection.
The free samples are stored in a roadmap (a graph structures), in which a path can be searched using standard graph-search methods.
A path in the roadmap then represents a motion in the workspace.

@@

The kinematic chain representaion used to model ligand flexibility allows the sampling-based planners to generate all possible conformations.
However, not all of them are feasible and it is therefore necessary to verify feasibility of the given conformation based on energy, which is time consuming.
Moreover, the aditional degrees of freedom required to model the flexiblity increases dimension of the configuration space.
RRT-based planners are sensitive to the employed metric, that should consider also the flexibility-related DOFs.

A RRT-based method for computing exit pathways for a small flexible molecule was presented in~\cite{cortes2010simulating}.
To cope with the many-DOF ligands, the RRT-ML variant~\cite{cortes2007mlrrt} was employed.
In RRT-ML expands the tree primarily using those DOF that are essential for achieving motion of the ligand (i.e., rotation
and translation) and it employs the other DOFs (i.e., that are responsible for conformation change) if they hinder the growth of the tree.
The approach~\cite{cortes2010simulating} aims to find any exit pathway for a given ligand, which requires
to search the whole configuration space of the ligand/protein complex.
This might be a possible disadvantage, as the proteins are dense structures with relatively narrow void space, which has the
properties of the narrow passage~\cite{hannaWIS}.
A narrow passage is a region in the configuration space whose removal changes the connectivy of the free space.
Narrow passages have smaller volume than other areas in it is therefore diffuclt to sample them dense enough using
uniform distribution.
Although there are usually multiple pathways from a given active site, not all of them are traversable by a given ligand.
Many iterations are needed to find a solution, which increases the computational time.

In this paper, we propose to employ RRT to compute the trajectory for a given ligand in a given tunnel.
We assume that the tunnel is computed using the standard tunnel detection tool and is represented using a sequence of spheres.
To guide the configuration tree through the configuration space, the random samples are generated gradually using the spheres of the tunnel.
As the tunnels are usually computed on protein models without ligands, they are usually very narrow and a ligand can barely traverse them.
It can be assume that both the protein and the ligand adapt to each other and therefore.
To enable the planner find a solution, the radii of the ligand, and possibly also the radii of the protein atoms, are down-scaled
by a given factor.

This brings two important advantages: the volume of the configuration space to be searched is decreased, as not the whole
protein/ligand complex, but only its subspace need to be searched.
Consequently, the relative volume of the corresponding free region of the configuration space is increased, which leads to easier motion planning.
The proposed approach allows the researches to analyze a given tunnel for various ligands without necessity to compute the tunnel
again for each ligand.

@@

Proteins and ligands are represented by the hard sphere model, where the radius of each sphere (atom) is given by its van der Waals radius.
%Let $\SS \subset \mathbf{R}^3$ denote the union of all spheres, i.e., the geometry of the protein.
%We further assume that a subset of atoms located at the surface, are known. These atoms, reffered to as {\sl surface atoms} in this paper,
%   can be detected as atoms located at $\alpha$-shape of the molecule.
%The tunnels need to be found for a spherical probe $\Sprobe$ of radius $\probe$.
The flexibility of a ligand is modeled using the set $\L$ of its conformations.
The conformations can be prepared e.g., considering the potential energy using tools like Rosetta.
We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.

@@

A protein tunnel is described by a sequence of collision-free spheres $T=( (t_i, r_i) )$, $i=1,\ldots,n$, where $n$ denotes the number of spheres,
$t_i=(x_i,y_i,z_i)$ is their 3D position and $r_i$ denotes the maximum collision-free radius of a sphere centered at $t_i$. 
The tunnels can be found by tunnel detection tools like CAVER 3.0~\cite{caver3}.

@@

The proteins are dense structures with narrow tunnels.
Depending on the protein, the tunnel bottlenecks can be even less then $1$~\AA, which is too narrow for ligands with more than 2 atoms.
For the purpose of traversability study, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
A discrete set of scales is used, i.e., $s \in \{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale and $\sdelta$ is the minimal difference between two scales.
This scaling technique is used also in other tools like MoMaLigPath~\cite{cortes2005path}.

A configuration of the pseudoflexible ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$  is described
by the 3D position $(x,y,z)$ of the reference point of the ligand (e.g., it's geometric center), rotation around $x$, $y$ and $z$ axes,
index of the conformation $l\in \L$ and the actual scale $s$.
All possible configurations form the configuration space $\C$. 
A configuration is collision-free if there is no intersection between the protein atoms and ligand atoms scaled by $s$.
%The set of all feasible configurations at scale $s$ is denoted $\CFD \subseteq \C$.

@@

The task of the traversability study is to determine if a given ligand can traverse  a given tunnel in the protein.
The spheres representing the tunnel are used to generate the random samples for the proposed RRT-based planner.
%We propose a RRT-based method to compute the trajectories of the ligand.
%As the tunnels are usually computed on proteins without ligands, their are usually very narrow and the classic RRT would
%suffer from the well known narrow passage problem~\cite{hannaWIS}.
%To cope with the narrow passage problem, the random samples are not generating from the whole configuration space.
%Instead, they are generated using the spheres of the tunnel.
The main loop of the method is described in Alg.~\ref{alg::main}.
In each iteration, a random sample $\qrand$ is generated, its nearest node $\qnear\in\T$ in the tree is found.
The nearest-neighbor search between the $\qrand$ and the tree is realized using the weighted 6D Euclidean metric considering
both 3D rotation and 3D translation.

To guide thee growth of the tree through the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
The virtual goal $v$ is the index of a sphere, around which the random samples are generated with probability $\gb$.
%The random samples are not generated only from whole $\C$, as in the original RRT, but the sampling
%is guided using moving virtual goal~\cite{vonasek2009rrt}.
After the tree reaches the sphere $t_v \in T$, i.e., the distance of the tree to the center of $t_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
Setting the virtual goal to this successor allows the tree to ``skip'' such parts of the tunnels that are
not traversible or reachable by the ligand.
The algorithm terminates after the predefined number of planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = N$.

The random samples $\qrand$ are generated around the virtual goal.
The translation of $\qrand$ is generated from Normal distribution centered at $t_v$ with radius $R_v$.
The radius $R_v$ is set as the absolute distance from the tunnel centerline, e.g. $R_v=3\AA$.
Alternatively, $R_v$ can be set to $R_v=r_v$ according the radius of the tunnel at the virtual goal.

\linesnumbered
\begin{algorithm}[h]
%\setstretch{\straa}
\caption{\label{alg::main}main loop of the RRT planner}
\KwIn{
    tunnel $T=(t_i, r_i), n=1,\ldots,N$ represented by $N$ spheres,
    initial configuration $\qinit$,
    goal configuration $\qgoal$,
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$ and $\sdelta$
}
\KwOut{
    configuration tree $\T$\;
}
\hrule
$v = 0$; // index of the virtual goal\\
$iteration = 0$\;
\While{$iteration < \Imax$ {\bf and}  $v < N$}{
    \eIf{$rand() < \gb$}{
        $\qrand$ = random sample around actual virtual goal $t_n \in T$\;
    }{
        $\qrand$ = random sample from $\C$\;
    }
    $\qnear$ = nearest node in $\T$ towards $\qrand$\;

    expand($\qnear$)\;

   % \For{$q \in S$}{
    %    $\T$.addNode($q$)\;
    %    $\T$.addEdge($\qnear, q$)\;
    %}

    \For{$i= N-1,N-2,\ldots,v+1,v$}{ \nllabel{alg::main:a}
        $d$ = nearest node in the tree towards sphere $t_i$\;
        \If{$d < \dt $}{
            $v = i+1$; // new virtual goal found\;
            {\bf  break}\;
        }
    } \nllabel{alg::main:b}
    $iteration = iteration+1$\;
}
\return $\T$\;
\end{algorithm}

The core of the proposed method is the expansion procedure (Alg.~\ref{alg::expand}) which generates new collision-free nodes around $\qnear\in\T$.
For each ligand conformation $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal
scale.
First, the maximal scale $\smax$ is tested and $m$ random samples are generated around $\qrand$ and tested for collision.
The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
If none of the tested samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the searches continues
until a free collision-free samples is found or until the mimimal reduced-scale $\smin$ is reached.

@@

The 3D position of $q$ is generated using Normal distribution $N(0,\sigma^2)$ around $\qnear$, where $\sigma^2$ is the predefined
parameter.
The rotation part of $q$ is generated using $U(0,2\pi)$.
The expansion begins with the largest scale $s$ and if less then $m$ collision-free samples are generated
with $s$, the next lower $s$ is used.
The expansion step is repeated for each conformation $l \in L$.

@@

\red{Adam: }
As described above, our approach enables to estimate the probability that a ligand given a minimum scale passes through a tunnel.
On the other hand, the probability does not describe any details about the ligand's passage, e.g., how hard it is to pass some part of the tunnel, which trajectory is the most used, etc.
Moreover, it is not feasible to directly visualize and explore the resulting trajectories due to the visual clutter (see Fig.~\ref{fig:trajectories} left) caused by their high number and close proximity.
Therefore, we propose to extract and visualize the complex information about the ligand's passage through a tunnel.
In this manner, we aim to further support the decision process of a chemist about a tunnel.
In order to do that, we employ an ensemble of resulting trajectories $T$, and we compute and visualize the following tunnel properties:
\begin{itemize}
  \item \textbf{Accessibility} --- The probability of the ligand that it reaches a tunnel's part.
We employ the tunnel representation from CAVER 3.0 and we define a tunnel part as a sphere $P = (x, r)$ where $x \in \mathbb{R}^3$ is sphere center and $r \in \mathbb{R}$ is radius.
Then for each trajectory $t$ we define its ending tunnel part $P_t^e$ as a tunnel part whose center is closest to the end position of $t$.
Finally, the accessibility of $P$ is defined as a ratio of trajectories that passed $P$ or ended in $P$ over all trajectories.
  \item \textbf{Ligand scale} --- The scale of the ligand when it passes a tunnel's part.
For the trajectory $t$ the ligand scale $s(P_t)$ in the tunnel part $P_t$ is defined as a scale of a such sample of $t$ whose position is the closest to the center of $P_t$.
Then, the average ligand scale in a tunnel part is defined as $s(P) = \frac{1}{|T|} \sum_{t \in T} s(P_t)$.
Similarly, minimum and maximum ligand scales can be defined.
  \item \textbf{Throughput} --- The probability of the ligand that it passes through a tunnel's part.
The througput of the tunnel part $P$ is defined as a ratio of trajectories that passed $P$ over all trajectories that reached $P$.
\end{itemize}
In the visualization, these properties are mapped onto a tunnel's surface (see Fig.~\ref{fig:properties}) so that they can be visually correlated with, e.g., the tunnel's shape.
More precisely, a point $x$ on the tunnel surface is colored according to the property value in its nearest tunnel part $P$, i.e., such part $P$ whose center is the closest to $x$ among all tunnel parts.

\begin{figure}
\centering
\begin{tabular}{ccc}
\includegraphics[width=0.3\textwidth]{fig/accessibility}
\quad & \quad
\includegraphics[width=0.3\textwidth]{fig/ligand-scale}
\quad & \quad
\includegraphics[width=0.3\textwidth]{fig/throughput}
\end{tabular}
\caption{Tunnel properties mapped onto its surface.
Left: Accessibility --- green denotes parts that were accesed by the ligand easily while red denotes parts that were accessible only in \tylde 10\% of computations.
Middle: Ligand scale --- the average scale of the ligand was \tylde 50\% in orange parts, \tylde 60\% in yellow parts, \tylde 80\% in yellow to green parts and \tylde 100\% in green parts.
Right: Throughput --- green denotes parts that are highly probable to be passed by the ligand while red denotes those that were hard to pass.}
\label{fig:properties}
\end{figure}

@@

guiding, parametes, different sampling, flexibility

>>>>

\def\rv{T_w}

@@

The widely used tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3} or 
a 3D grid~\cite{sehnal2013mole,petrek2006caver}.
These methods compute the tunnels assuming a spherical ligand (probe), and the resulting tunnels are represented as a sequence of spheres.
Bio-chemists then decide if such a tunnel can be used to transport a given ligand mainly based on the tunnel lengths and bottleneck (i.e.,
        the radius of the smallest sphere that can traverse the whole tunnel).
Typical ligands are however non-spherical and it is therefore difficult to estimate their traversability in tunnels computed for a spherical probe.
The decision based only on the spherical tunnels requires previous expertise in the domain and yet it may be imprecise.

To provide chemists more insight into the behavior of non-spherical ligands, it is necessary to compute the trajectories of the ligand considering it's shape and possibly also conformation changes, which can be formulated as a path planning problem in the a high-dimensional configuration space.

@@

This requires to compute trajectories of the non-spherical ligand in the tunnels, which can be formulated as a path planning problem in a high-dimensional configuration space $\C$.
The configuration space is formed by all possible configurations of the ligand in the tunnel, i.e., considering its rotation, translation and possibly also other degrees of freedom responsible for the conformation changes.
The dimension of the configuration space is given by the DOF of the ligand, i.e., 6D for a rigid ligand and 6D + $n$ for a flexible
ligand with $n$ DOFs.
Sampling-based motion planning methods~\cite{Lav06} can be used to search this high-dimensional configuration space.
The idea of sampling-based motion planning is to randomly sample $\C$ and classify the samples as free or non-free using collision detection.
The free samples are stored in a roadmap (a graph structures), in which a path can be searched using standard graph-search methods.
%A path in the roadmap then represents a motion in the workspace.

@@

A well known issue of sampling-based planners is the narrow passage problem~\cite{hannaWIS}.
A narrow passage is a region in the configuration space whose removal changes the connectivity of the free space.
Narrow passages have smaller volume than other regions and it is therefore difficult to sample them dense enough using uniform distribution, that is used in the basic sampling-based planners.
The presence of narrow passages in the configuration space, especially if they contain part of the solution, requires many iterations
in order to put enough samples there and it consequently increases the planning time.
The protein tunnels have the properties of the narrow passages. 
A possible approach to cope with the narrow passage property of the protein tunnels is to use the information
about their 3D position, which is represented by the tunnel centerline.
The centerline can be used to guide the growth of the RRT tree through the configuration space~\cite{vonasek2009rrt,denny2016dynamic}.

The kinematic chain representation used to model ligand flexibility allows the sampling-based planners to generate all possible conformations.
However, not all of them are feasible and it is therefore necessary to verify feasibility of the given conformation based on energy, which is time consuming.
Moreover, the additional degrees of freedom required to model the flexibility increases dimension of the configuration space.
RRT-based planners are sensitive to the employed metric, that should consider also the flexibility-related DOFs.

A RRT-based method for computing exit pathways for a small flexible molecule was presented in~\cite{cortes2010simulating}.
To cope with the many-DOF ligands, the RRT-ML variant~\cite{cortes2007mlrrt} was employed.
In RRT-ML expands the tree primarily using those DOF that are essential for achieving motion of the ligand (i.e., rotation
and translation) and it employs the other DOFs (i.e., that are responsible for conformation change) if they hinder the growth of the tree.
The approach~\cite{cortes2010simulating} may however suffer from the narrow passage problem, as it aims to find any exit pathway for a given ligand, which requires to search the whole configuration space of the ligand/protein complex.
Although there are usually multiple pathways from a given active site, not all of them are traversable by a given ligand.
Many iterations are needed to find a solution, which increases the computational time.

In this paper, we propose to analyze the traversability separately for each tunnel.
This can bring several advantages for the biochemists.
The tunnels in a protein being studied can be computed using standard tunnel detection tools for spherical probes and assessed
according their length, curvature, bottleneck and biochemical properties (e.g. partial charge).
These characteristics are already used in many research studies.
Selected promising tunnels can be further analyzed for a given ligand to determine the traversability.
It is not necessary to first run time consuming MD simulations with both protein and ligand being studied, but it is suitable
to detected tunnels in the protein only.

The proposed work employs the RRT planning principle with a three extensions: 
a) the configuration space is not sampled uniformly, but using a guided sampling along the tunnel centerline,
b) the radii of the ligand atoms are decreased, and 
c) ligand flexibility of modeled using a set of predefined energy-suitable conformations.
The first two extensions can help to cope with the narrow passage problem and the third modification
is introduced in order to decrease the dimension of the configuration space and the sensitivity to the employed
metric.

\red{presunu jinam}
This brings two important advantages: the volume of the configuration space to be searched is decreased, as not the whole
protein/ligand complex, but only its subspace need to be searched.
Consequently, the relative volume of the corresponding free region of the configuration space is increased, which leads to easier motion planning.
The proposed approach allows the researches to analyze a given tunnel for various ligands without necessity to compute the tunnel
again for each ligand.

@@

Proteins and ligands are represented by the hard sphere model, where the radius of each sphere (atom) is given by its van der Waals radius.
%Let $\SS \subset \mathbf{R}^3$ denote the union of all spheres, i.e., the geometry of the protein.
%We further assume that a subset of atoms located at the surface, are known. These atoms, reffered to as {\sl surface atoms} in this paper,
%   can be detected as atoms located at $\alpha$-shape of the molecule.
%The tunnels need to be found for a spherical probe $\Sprobe$ of radius $\probe$.
The flexibility of a ligand is modeled using the set $\L$ of its conformations.
The conformations can be prepared e.g., considering the potential energy.
We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.

@@

A protein tunnel is described by a sequence of collision-free spheres 
$T=( (t_1, r_1),\ldots,(t_n,r_n) )$, where $n$ denotes the number of spheres,
$t_i=(x_i,y_i,z_i)$ is their 3D position and $r_i$ denotes the maximum collision-free radius of a sphere centered at $t_i$. 
The tunnels can be found by tunnel detection tools like CAVER 3.0~\cite{caver3}.

@@

The proteins are dense structures with narrow tunnels.
Depending on the protein, the tunnel bottlenecks can be even less than $1$~\AA, which is too narrow for ligands with more than 2 atoms.
For the purpose of traversability study, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
A discrete set of scales is used, i.e., $s \in \{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale and $\sdelta$ is the minimal difference between two scales.
This scaling technique is used also in other tools like MoMaLigPath~\cite{cortes2005path}.

A configuration of the pseudoflexible ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$  is described
by the 3D position $(x,y,z)$ of the reference point of the ligand (e.g., it's geometric center), rotation around $x$, $y$ and $z$ axes,
index of the conformation $l\in \L$ and the actual scale $s$.
All possible configurations form the configuration space $\C$. 
A configuration is collision-free if none of the ligand atoms scaled by $s$ and placed at the
position defined by $q$ collide with the protein atoms.
%The set of all feasible configurations at scale $s$ is denoted $\CFD \subseteq \C$.

@@

The traversability study proposed in this paper is achieved by computing many trajectories for the given ligand and tunnel, which
is solved using a novel RRT-based planner.
The main principle of the RRT planner is extended by guided sampling, scaling of the ligands atoms and using predefined set of ligand
conformations.
The main loop of the method is described in Alg.~\ref{alg::main}.
%The spheres representing the tunnel are used to generate the random samples for the proposed RRT-based planner.
%We propose a RRT-based method to compute the trajectories of the ligand.
%As the tunnels are usually computed on proteins without ligands, their are usually very narrow and the classic RRT would
%suffer from the well known narrow passage problem~\cite{hannaWIS}.
%To cope with the narrow passage problem, the random samples are not generating from the whole configuration space.
%Instead, they are generated using the spheres of the tunnel.
In each iteration, a random sample $\qrand$ is generated, its nearest node $\qnear\in\T$ in the tree is found.
The nearest-neighbor search between the $\qrand$ and the tree is realized using the weighted 6D Euclidean metric considering
both 3D rotation and 3D translation.

To guide tree growth of the tree through the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
The virtual goal $v, 1\le v \le n$, is the index of a tunnel sphere.
The random samples $\qrand$ are generated with probability $1-\gb$ from $\C$, otherwise they are generated
around the virtual goal sphere $v$.
%around which the random samples are generated with probability $\gb$.
%The random samples are not generated only from whole $\C$, as in the original RRT, but the sampling
%is guided using moving virtual goal~\cite{vonasek2009rrt}.
After the tree reaches the sphere $t_v \in T$, i.e., the distance of the tree to the center of $t_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are
not traversable or reachable by the ligand.
This is necessary in the dense protein structures, where it is not always possible to exactly follow the tunnel.
The algorithm terminates after the predefined number of planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = n$.

To generate the samples $\qrand$ around the virtual goal $v$, the translation part $(x,y,z)$ of $\qrand$ is generated
from $N(t_v,\Sigma)$, where $\Sigma$ is diagonal matrix with diagonal entries equal to a parameter $\rv$, and the rotational
part of $\qrand$ is generated using techniques described in~\cite{kuffnerES}.
The other two parameters (ligand index $l$ and scale $s$) of $\qrand$ can be left zero, as these are not used in the employed
metric for nearest-neighbor search.

\linesnumbered
\begin{algorithm}[h]
%\setstretch{\straa}
\caption{\label{alg::main}main loop of the RRT planner}
\KwIn{
    tunnel spheres $t_i$ and their radii $r_i$, $i=1,\ldots,n$, 
    initial configuration $\qinit$,
    goal configuration $\qgoal$,
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$ and $\sdelta$
}
\KwOut{
    configuration tree $\T$\;
}
\hrule
$v = 1$; // index of the virtual goal\\
$iteration = 0$\;
\While{$iteration < \Imax$ {\bf and}  $v < n$}{
    \eIf{$rand() < \gb$}{
        $\qrand$ = random sample around actual virtual goal $t_n \in T$\;
    }{
        $\qrand$ = random sample from $\C$\;
    }
    $\qnear$ = nearest node in $\T$ towards $\qrand$\;
    expand($\qnear$)\;
    \For{$i= n-1,n-2,\ldots,v+1,v$}{ \nllabel{alg::main:a}
        $d$ = nearest node in the tree towards sphere $t_i$\;
        \If{$d < \dt $}{
            $v = i+1$; // new virtual goal found\;
            {\bf  break}\;
        }
    } \nllabel{alg::main:b}
    $iteration = iteration+1$\;
}
\return $\T$\;
\end{algorithm}

The core of the proposed method is the expansion procedure (Alg.~\ref{alg::expand}) which generates new collision-free nodes around $\qnear\in\T$.
For each ligand conformation $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal
scale.
First, the maximal scale $\smax$ is tested and $m$ random samples are generated around $\qrand$ and tested for collision.
The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
If none of the tested samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the searches continues
until a free collision-free samples is found or until the minimal reduced-scale $\smin$ is reached.

@@

The random samples $q$ are generated similarly as in the case of $\qrand$ samples, only their translation
part is generated around $\qnear$.
The expansion begins with the largest scale $s$ and if less then $m$ collision-free samples are generated with $s$, the next lower $s$ is used.
The expansion step is repeated for each conformation $l \in L$.

@@

\red{Adam: }
As described above, our approach enables to estimate the probability that a ligand given a minimum scale passes through a tunnel.
On the other hand, the probability does not describe any details about the ligand's passage, e.g., how hard it is to pass some part of the tunnel, which trajectory is the most used, etc.
Moreover, it is not feasible to directly visualize and explore the resulting trajectories due to the visual clutter (see Fig.~\ref{fig:trajectories} left) caused by their high number and close proximity.
Therefore, we propose to extract and visualize the complex information about the ligand's passage through a tunnel.
In this manner, we aim to further support the decision process of a chemist about a tunnel.
In order to do that, we employ an ensemble of resulting trajectories $T$, and we compute and visualize the following tunnel properties:
\begin{itemize}
  \item \textbf{Accessibility} --- The probability of the ligand that it reaches a tunnel's part.
We employ the tunnel representation from CAVER 3.0 and we define a tunnel part as a sphere $P = (x, r)$ where $x \in \mathbb{R}^3$ is sphere center and $r \in \mathbb{R}$ is radius.
Then for each trajectory $t$ we define its ending tunnel part $P_t^e$ as a tunnel part whose center is closest to the end position of $t$.
Finally, the accessibility of $P$ is defined as a ratio of trajectories that passed $P$ or ended in $P$ over all trajectories.
  \item \textbf{Ligand scale} --- The scale of the ligand when it passes a tunnel's part.
For the trajectory $t$ the ligand scale $s(P_t)$ in the tunnel part $P_t$ is defined as a scale of a such sample of $t$ whose position is the closest to the center of $P_t$.
Then, the average ligand scale in a tunnel part is defined as $s(P) = \frac{1}{|T|} \sum_{t \in T} s(P_t)$.
Similarly, minimum and maximum ligand scales can be defined.
  \item \textbf{Throughput} --- The probability of the ligand that it passes through a tunnel's part.
The throughput of the tunnel part $P$ is defined as a ratio of trajectories that passed $P$ over all trajectories that reached $P$.
\end{itemize}
In the visualization, these properties are mapped onto a tunnel's surface (see Fig.~\ref{fig:properties}) so that they can be visually correlated with, e.g., the tunnel's shape.
More precisely, a point $x$ on the tunnel surface is colored according to the property value in its nearest tunnel part $P$, i.e., such part $P$ whose center is the closest to $x$ among all tunnel parts.

\begin{figure}
\centering
\begin{tabular}{ccc}
\includegraphics[width=0.3\textwidth]{fig/accessibility}
\quad & \quad
\includegraphics[width=0.3\textwidth]{fig/ligand-scale}
\quad & \quad
\includegraphics[width=0.3\textwidth]{fig/throughput}
\end{tabular}
\caption{Tunnel properties mapped onto its surface.
Left: Accessibility --- green denotes parts that were accessed by the ligand easily while red denotes parts that were accessible only in \tylde 10\% of computations.
Middle: Ligand scale --- the average scale of the ligand was \tylde 50\% in orange parts, \tylde 60\% in yellow parts, \tylde 80\% in yellow to green parts and \tylde 100\% in green parts.
Right: Throughput --- green denotes parts that are highly probable to be passed by the ligand while red denotes those that were hard to pass.}
\label{fig:properties}
\end{figure}

@@

guiding, parameters, different sampling, flexibility
